---
title: "EDA of White Wine Quality Data"
author: "Sohan Aryal"
date: "7/15/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Loading Libraries
```{r, results = "hide"}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(psych)
library(corrplot)

```

#### Loading Data
```{r}
wine <- read.csv2("winequality-white.csv")
glimpse(wine)
```

Since all variable except quality are as character we need to convert them from character into numerical.

```{r}
wine <- data.frame(lapply(wine, function(x) as.numeric(as.character(x))))

```

Let's see the summary of each variable in dataset.

```{r}
summary(wine)
```

#### Wine Quality Distribution
Our main variable of interest in this dataset is wine quality. Let's look at the summary of wine quality in the following table and histogram.

```{r, warning=FALSE}
print("Number of wines with particular rating:")
table(wine$quality)

ggplot(aes(x = quality), data = wine) + 
  geom_histogram(stat = "count") +
  xlab("Quality of White Wines") +
  ylab("Number of White Wines")
```

Quality of white wine has near-normal distribution as seen from the plot. Most of the wines are rated 5 to 7, while the very few are rated [3,4] and [8,9]. There is no wine with rating 1, 2 and 10.


#### Univariate Analysis

```{r Univariate Plots, fig.height = 8, fig.width = 13, message=FALSE}
p1 <- ggplot(aes(x=fixed.acidity),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Fixed acidity', y = NULL)

p12 <- ggplot(aes(x=fixed.acidity),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Fixed acidity', y = NULL)

p2 <- ggplot(aes(x=volatile.acidity),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Volatile acidity', y = NULL)

p22 <- ggplot(aes(x=volatile.acidity),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Volatile acidity', y = NULL)

p3 <- ggplot(aes(x=citric.acid),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Citric acid', y = NULL)

p32 <- ggplot(aes(x=citric.acid),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Citric acid', y = NULL)

p4 <- ggplot(aes(x=residual.sugar),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Residual sugar', y = NULL)

p42 <- ggplot(aes(x=residual.sugar),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Residual sugar', y = NULL)

p5 <- ggplot(aes(x=chlorides),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Chlorides', y = NULL)

p52 <- ggplot(aes(x=chlorides),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Chlorides', y = NULL)

p6 <- ggplot(aes(x=free.sulfur.dioxide),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Free sulfur dioxide', y = NULL)

p62 <- ggplot(aes(x=free.sulfur.dioxide),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Free sulfur dioxide', y = NULL)

p7 <- ggplot(aes(x=total.sulfur.dioxide),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Total sulfur dioxide', y = NULL)

p72 <- ggplot(aes(x=total.sulfur.dioxide),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Total sulfur dioxide', y = NULL)

p8 <- ggplot(aes(x=density),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Density', y = NULL)

p82 <- ggplot(aes(x=density),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Density', y = NULL)

p9 <- ggplot(aes(x=pH),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'pH levels', y = NULL)

p92 <- ggplot(aes(x=pH),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'pH levels', y = NULL)

p10 <- ggplot(aes(x=sulphates),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Sulphates', y = NULL)

p102 <- ggplot(aes(x=sulphates),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Sulphates', y = NULL)

p11 <- ggplot(aes(x=alcohol),data = wine) +
  geom_histogram(color = "black", fill = "gray") +
  labs(x = 'Alcohol', y = NULL)

p112 <- ggplot(aes(x=alcohol),data = wine) +
  geom_boxplot(color = "black", fill = "gray") +
  labs(x = 'Alcohol', y = NULL)

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, ncol = 3, top = "Univariate Plots (Histogram)")

grid.arrange(p12, p22, p32, p42, p52, p62, p72, p82, p92, p102, p112, ncol = 2, top ="Box-Plot")
```


Variables like fixed acidity, volatile acidity and citric acid have outliers; which when removed their distribution looks symmetric. Some variables like residual sugar doesn't become symmetric even when their outliers are removed. Mostly outliers of these variables are on the larger values. Alcohol doesn't have outliers but have some irregular type of distribution.


#### Bivariate Plots

Let's do the correlation plot to see how associated variables are.

```{r Bivariate_Plots, fig.height=10, fig.width=10}

wq.corr <- cor(wine)
corrplot(wq.corr, method='color', type = 'upper', 
         tl.col = "black", addCoef.col='blue')
```

From correlation matrix, we see how quality of wine is related with other chemical properties of wine.
* Quality and alcohol have moderate correlation (0.44), it is highest amongst all other chemical properties.
* Quality and density has lower negative correlation (-0.31).
* Quality has very low negative correlation with total sulfur dioxide, chlorides and volatile acidity.
* Quality has very weak negative correlation with residual sugar(-0.1), citric acid(-0.01) and fixed acidity(-0.11).
* Quality has very weak positive correlation with pH(0.1), sulphates(0.05) and free sulphur dioxide(0.01).


* Alcohol has strong negative correlation (-0.78) with the density of wine.
* Density is strongly correlated (0.84) with residual sugar quantity.
* Alcohol has moderate negative correlations with Residual sugar(-0.45), Total SO2 (-0.45) and Chlorides (-0.36).


Overall it was seen that main taste qualities of wine, such as acidity and sweetness have a low effect on quality. The highest positive correlation of quality rating appears to be with alcohol. Density is a variable which seems to be affected by other variables the most. In fact strongest relationship in the given sample is with Density and Alcohol.


#### Data Preparation

From above plot, we have seen that most outliers are on higher end of values, so we consider those observation as outlier which are above $$Q_3 + 1.5 * IQR$$.

```{r}

df <- wine

limit <- rep(0,11)

for (i in 1:11){
  q3 <- quantile(df[,i], 0.75)
  iqr <- IQR(df[,i], 0.75)
  
  limit[i] <- q3 + 1.5*iqr
}

df_index <- matrix(0, 4898,11)



for (i in 1:4898)
  for (j in 1:11)
    if (df[i,j] > limit[j])
      df_index[i,j] <- 1


dfIndex <- base::apply(df_index, 1, sum)


df2 <- df[dfIndex == 0, ]

indexes = sample(1:nrow(df2), size=0.7*nrow(df2))
train <- df2[indexes,]
test <- df2[-indexes,]
```


Above procedure reduces the data size from 4899 to 4074. Data is randomly divided into Training data and Test Data in ratio of 7:3.


#### Regression Model

Now we run linear regression to find out which variables are significant. We go on improving our model by omitting insignificant ones by checking improvement in R-square value.

Now we will create the model to predict quality of wines. We use the train data set.

```{r}
model1 <- lm(quality ~ ., data = train)
summary(model1)
```

Citric acid and Total sulfur dioxide are not significant in this model with adjusted R-square of 0.2407.
Now we create model without these variables.

```{r}
model2 <- lm(quality ~ . -citric.acid -total.sulfur.dioxide, data = train)
summary(model2)
```

It didn't quite improve the model. But all the variables are significant now.
We can see that predictors are continuous while quality is integer. Prediction made will not be integer, so it is difficult to explain the variation in quality using linear model. We can check the prediction by our model2.

```{r}
predictTest <- predict.lm(model2, newdata = test)
summary(predictTest)
```

The model2 didn't predict quality over 7.165. So now we will use different method for prediction.


# Logistic Regression

We will make a logistics regression model to predict the quality of wine using input variables.
First we will create a subset of wines rated 8 or above effectively making quality having two factors.
Quality rating > 7 : wine is "Good"
Quality rating <=7 : wine is "Not good"

```{r}

train$rating <- ifelse (as.integer(train$quality) > 7, 1, 0)

test$rating <- ifelse (as.integer(test$quality) > 7, 1, 0)

model3 = glm(rating ~.-quality ,data=train,family = "binomial")
summary(model3)
```

Significant variables are: fixed.acidity, residual.sugar and pH.
We will use this model to predict over testing set.

```{r}
prediction = predict(model3, newdata=test, type="response")
table(test$rating, prediction > 0.5)
```

Our model predicts that no wine is in the best rated wine group, but actually 54 of the wines are rated 8 or 9 in the test set. Our prediction is same as the baseline prediction. So, there is not much benefit from this model. Let's check what is the maximum value of prediction over test set.

```{r}
print("The maximum value of prediction over testing set is ")
round(max(prediction),3)
```

As the maximum value is less than 0.5, our model will always predict that wine quality will be less than 8.



### Conclusion

We used linear regression model and logistic regression model. Though we see some association from correlation plot and significant variables on these two models, it seems like wine quality is not well supported by its chemical properties. At each quality level variability of the predictors is high.